Supplmental Information : Understanding the nanoscale structure of hexagonal phase lyotropic liquid crystal membranes

### 2018-01-19 Ben_Manuscripts/structure_paper/Draft.tex
Added new figures
More analysis scripts and figures

@@

  \item What constitutes a pore, and how well-defined are the pore regions? \label{point:poredefinition}

@@

  When constructing an initial configuration, there are a number of variables
  which require careful consideration while placing monomers. The equilibrium
  configuration is sensitive to some while insensitive to others. The starting
  pore radius, defined as the distance of a chosen head group carbon from the
  pore's central axis, does not influence the equilibrium structure if one choses
  a reasonable value (See Supplemental Information). The pore radius is chosen to
  be 0.6 nm in our initial configurations because the pore size is estimated to
  be $\sim$1.2 nm \cite{zhou_supported_2005}. The initial distance between pores
  also has little effect on the the equilibrated structure (See supplemental
  information). However, one should not start them too close or there will be
  high energy repulsions during early equilibration. We chose an initial pore
  spacing of 4.5 nm, $\sim$10\% larger than the experimental value of 4.12 nm.
  The distance between layers, the rotation of the layers with respect to
  adjacent layers, and the number of monomers per layer do influence the
  equilibrium structure and require further justification for their choices.  We
  rely on experimental data to inform them.

@@

  \subsection{Determining Definition of Layers}

  We learned that layers are well-defined and persistent, answering
  (\ref{point:layers}). We established our conclusion by plotting the pair
  correlation function, $g(z)$, calculated between atoms along the length of the
  pores (Fig.~\ref{fig:zdf}). We measured $g(z)$ with respect to aromatic rings in
  the head groups and, separately, with respect to carbon atoms in the alkane
  chains. Using a Fourier transform of $g(z)$, we see that sandwiched
  configuration layers stack 4.39 \angstrom~apart while parallel displaced
  configuration layers stack 4.38 \angstrom~apart.

@@

%MRS4: can we come up with another name?  I still think disordered pores is a decent name, since there is no pi-stacking layering.  ``structures that form when layers are 5 AA apart'' is not so good.  or ``more disordered''?
%BJC3: agreed, I'll stick with disordered pores
%  The $z$-direction correlation functions show that layers in our model prefer
%  to stack further apart than 3.7 \angstrom~as suggested by experiment.
  When we build systems with layers stacked 5.0 \AA~apart and then let them
  equilibrate, we observe long-term stability of a qualitatively different
  configuration, suggesting that we have found another metastable free energy
  basin, further corroborating (\ref{point:metastable}). For reasons to become
  apparent later, we will refer to this class of systems as the disordered pore
  basin. Systems which we build with layers initially stacked 3.7 \AA~apart will
  be referred to as the ordered basin. We studied each basin in both the parallel
  displaced and sandwiched configurations.

%  Structural properties are different when layers are initially spaced 5 \AA~
%  apart. 
  Structural properties are different in the disordered pore basin.  In both
  configurations, we observe a decrease in pore spacing
  (Fig.~\ref{fig:p2p_disordered}) and a corresponding increase in the
  equilibrated distance between layers (Fig.~\ref{fig:dbwl_disordered}). The
  simulated X-ray diffraction patterns indicate further structural differences.
  In the parallel displaced configuration, almost all contrast between R-spots
  and R-alkanes fades (Fig.~\ref{fig:offset_disordered_xrd}). In the
  sandwiched configuration, R-spots is weakly present, but in different
  locations, showing higher intensity at the top and bottom of the pattern as
  well as at the intersection of R-alkanes with q$_z$ = 0 (Fig.
  ~\ref{fig:layered_disordered_xrd}).

@@

@@

  \subsection{Pore Structure Depends on Initial Configuration}  % Addresses pore definition

@@

  In all cases, the space in the pore region is filled with sodium ions and
  head groups. In the ordered pore systems, there is less density in the center of
  the pore indicating a more ordered pore structure. In both the sandwich and
  parallel displaced configurations, we see the density of head groups and sodium
  ions fall to less than 50\% of its maximum at r = 0
  (Fig.~\ref{fig:layered_density}). The situation is most pronounced in the
  sandwiched configuration where the maximum head group density occurs 0.44 nm
  from the pore center. The parallel displaced configuration reaches its maximum
  0.35 nm from the pore center. In contrast, both disordered pore systems show
  very little difference in density from its maximum. This implies a more uniform
  distribution of head groups within the pore center. 
  % High vacancy is less likely to be stable in the real system

  There is a partition between the hydrophobic and hydrophilic regions, however
  it is a gradient in composition, rather than an abrupt division. The system
  does not confine sodium ions and head groups to just within the pore region.
  Assuming a pore radius of 0.6 nm, we see in all cases, that 19\% of sodium ions
  exist outside the pore region (except sandwiched, ordered pore , where 16\%
  are outside the pore). Additionally, we see that in all cases, about 3\% of the
  plotted tail density is located within the pore region (except sandwiched,
  ordered pore, where 1.5\% are within the pore region). These observations bring
  into question how one should define a pore in these types of systems. One
  usually measures a membrane's pore radius based on the size of a molecule it
  can reject, however it is not clear where the edges of the pores are and what
  size molecule would fit through. We leave these investigations for a future
  study.

>>>>

  \begin{abstract}

  Nanostructured porous membranes made from the cross-linked hexagonal phase of
  self-assembled lyotropic liquid crystals (LLCs) are a promising material for
  selective separations. To date, their separation performance has only been
  characterized with membranes made from a small set of LLC monomers. There
  remains a large design space that we can likely exploit to design membranes for
  solute specific separations. In this work, we develop an atomistic molecular
  model of an LLC membrane. We show that our model is maximally consistent with
  experimental observations by comparing simulated X-ray Diffraction (XRD)
  patterns and calculated ionic conductivity to their experimental counterparts.
  We explore, in depth, the composition and structure of the nanopores in order
  to give insights that are not easily accessible to experimentalists. With the
  clear picture of the nanoscopic structure of these membranes gained from this
  work, we will be able to use the model to understand the mechanisms of small
  charged and uncharged solute transport within the nanopores, and eventually we
  will be able to screen large sets of modified monomers with their separation
  properties in mind.

  \end{abstract}

@@

  %\item What constitutes a pore, and how well-defined are the pore regions? \label{point:poredefinition}
  \item What constitutes a pore, and what does the pore architecture look like? \label{point:poredefinition}

@@

  When constructing an initial configuration, there are a number of variables
  which require careful consideration while placing monomers. The equilibrium
  configuration is sensitive to some while insensitive to others. The starting
  pore radius, defined as the distance of a chosen head group carbon from the
  pore's central axis, does not influence the equilibrium structure if one choses
  a reasonable value (See Supplemental Information). The pore radius is chosen to
  be 0.5 nm in our initial configurations. The initial distance between pores
  also has little effect on the the equilibrated structure (See supplemental
  information). However, one should not start them too close or there will be
  high energy repulsions during early equilibration. We chose an initial pore
  spacing of 4.5 nm, $\sim$10\% larger than the experimental value of 4.12 nm.
  The distance between layers, the rotation of the layers with respect to
  adjacent layers, and the number of monomers per layer do influence the
  equilibrium structure and require further justification for their choices.  We
  rely on experimental data to inform them.

@@

  \subsection{Layer partitioning}

  We learned that layers are well-defined and persistent, answering
  (\ref{point:layers}). We established our conclusion by plotting the one dimensional pair
  distribution function, $g(z)$, calculated between atoms along the length of the
  pores (Fig.~\ref{fig:zdf}). We measured $g(z)$ with respect to aromatic rings in
  the head groups and, separately, with respect to carbon atoms in the alkane
  chains. Using a Fourier transform of $g(z)$, we see that sandwiched
  configuration layers stack 4.39 \angstrom~apart while parallel displaced
  configuration layers stack 4.38 \angstrom~apart.

  Insert paragraph about 2D pair distribution function (yz cross section).

  We measured the correlation length, L, by fitting a function of the form:
  $$1 + Asin(Bx + C)e^{-z/L}$$ where z is the independent variable and all
  other variables are fitting parameters. The correlation length for each
  system is given in Table~\ref{fig:correlation_length}.

  \begin{table}[h]
  \centering
  \begin{tabular}{ccc}
  \toprule
  System & Layer Separation & Correlation Length \\
  \midrule
  Sandwiched & 4.39 \AA & 41 $\pm$ 3 \AA \\
  Parallel Displaced & 4.38 \AA & 86 $\pm$ 14 \AA \\
  Experiment* & 3.7 \AA & 32 $\pm$ 1 \AA \\  
  \bottomrule
  \end{tabular}
  \caption{*See supplemental information for derivation of experimental correlation length}~\label{fig:correlation_length}
  \end{table}

@@

%MRS4: can we come up with another name?  I still think disordered pores is a decent name, since there is no pi-stacking layering.  ``structures that form when layers are 5 AA apart'' is not so good.  or ``more disordered''?
%BJC3: agreed, I'll stick with disordered pores
%  The $z$-direction correlation functions show that layers in our model prefer
%  to stack further apart than 3.7 \angstrom~as suggested by experiment.
  When we build systems with layers stacked 5.0 \AA~apart and then let them
  equilibrate, we observe long-term stability of a qualitatively different
  configuration, suggesting that we have found another metastable free energy
  basin, further corroborating (\ref{point:metastable}). The basins are
  differentiated by their relative degrees of ordering resulting from different
  spatial arrangements.

  We expect that the head groups in more ordered systems will be predominantly
  coplanar with the xy plane. We defined a director vector, \textbf{n}, as the
  vector normal to the plane of the monomer's aromatic head group, then measured
  the distribution of angles between it and the z-axis. We used the resulting
  distibution of angles to calculate the nematic order parameter, S. The values
  of S for systems resulting from each initial configuration are presented in
  Table \ref{table:nematic}. The full distributions of angles between \textbf{n}
  and the z-axis are presented in the supplemental material. In general, systems
  started with layers spaced 3.7 \AA apart are more nematically ordered than
  those initially spaced 5 \AA apart. We will refer to systems built with layers
  initially spaced 3.7 \AA apart and 5.0 \AA apart as the ordered and disordered
  pore basins respectively. We studied each basin in both the parallel displaced
  and sandwiched configurations.

  \begin{table}[h]
  \centering
  \begin{tabular}{cc}
  \toprule
  System & Nematic Order Parameter (S) \\
  \midrule
  Sandwiched, Ordered & 0.659 $\pm$ 0.003 \\
  Parallel Displaced, Ordered & 0.601 $\pm$ 0.003 \\
  Sandwiched, Disordered & 0.565 $\pm$ 0.002 \\
  Parallel Displaced, Disordered & 0.503 $\pm$ 0.002 \\
  \bottomrule
  \end{tabular}
  \caption{The nematic order parameter is higher when layers are initially spaced
  3.7 \AA apart. A perfectly stacked system with aromatic rings coplanar with the
  xy plane would have a nematic order parameter of 1, while isotropically aligned
  director vectors would have a nematic order parameter of 0.}~\label{table:nematic}
  \end{table}

  We observe more evidence of disorder in the disordered pore basin when we 
  plot the two dimensional pair distribution function between the centers of 
  mass of monomer head groups. ADD MORE ANALYSIS ONCE CODE IS WORKING HOW I WANT IT

  \begin{figure}
  \begin{subfigure}{1\linewidth}
        \centering
        \begin{subfigure}{0.45\linewidth}
                \centering
                \includegraphics[width=\linewidth]{layered_xy_rings.png}
                \caption{}~\label{fig:sandwich_xy}
        \end{subfigure}%
        \begin{subfigure}{0.45\linewidth}
                \centering
                \includegraphics[width=\linewidth]{offset_xy_rings.png}
                \caption{}~\label{fig:offset_xy}
        \end{subfigure}
        \begin{subfigure}{0.45\linewidth}
                \centering
                \includegraphics[width=\linewidth]{disorder_sandwich_xy_rings.png}
                \caption{}~\label{fig:disorder_sandwich_xy}
        \end{subfigure}%
        \begin{subfigure}{0.45\linewidth}
                \centering
                \includegraphics[width=\linewidth]{disorder_offset_xy_rings.png}
                \caption{}~\label{fig:disorder_offset_xy}
        \end{subfigure}
  \end{subfigure}
  \caption{}\label{fig:xy_correlation}
  \end{figure}

  The simulated X-ray diffraction patterns indicate further structural
  differences. In the parallel displaced configuration, almost all contrast
  between R-spots and R-alkanes fades (Fig.~\ref{fig:offset_disordered_xrd}). In
  the sandwiched configuration, R-spots is weakly present, but in different
  locations, showing higher intensity at the top and bottom of the pattern as
  well as at the intersection of R-alkanes with q$_z$ = 0 (Fig.
  ~\ref{fig:layered_disordered_xrd}).

@@

  The equilibrated pore spacing and distance between layers are different in
  the disordered pore basin. For both head group stacking arrangements, we
  observe a decrease in pore spacing (Fig.~\ref{fig:p2p_disordered}) and a
  corresponding increase in the equilibrated distance between layers
  (Fig.~\ref{fig:dbwl_disordered}).

@@

  \subsection{Pore Composition Depends on Initial Configuration}

@@

  In all cases, the space in the pore region is filled with sodium ions and
  head groups. In the ordered pore systems, there is less density in the center of
  the pore indicating a more ordered pore structure. In both the sandwiched and
  parallel displaced configurations, we see the density of head groups and sodium
  ions fall to less than 50\% of its maximum at r = 0
  (Fig.~\ref{fig:layered_density}). The situation is most pronounced in the
  sandwiched configuration where the maximum head group density occurs 0.44 nm
  from the pore center. The parallel displaced configuration reaches its maximum
  0.35 nm from the pore center. In contrast, both disordered pore systems show
  very little difference in density from its maximum. This implies a more uniform
  distribution of head groups within the pore center. 
  % High vacancy is less likely to be stable in the real system

  There is a partition between the hydrophobic and hydrophilic regions, however
  it is a gradient in composition, rather than an abrupt division. The system
  does not confine sodium ions and head groups to just within the pore region.
  Assuming a pore radius of 0.6 nm, we see in all cases, that 19\% of sodium ions
  exist outside the pore region (except sandwiched, ordered pore, where 16\%
  are outside the pore). Additionally, we see that in all cases, about 3\% of the
  plotted tail density is located within the pore region (except sandwiched,
  ordered pore, where 1.5\% are within the pore region). These observations bring
  into question how one should define a pore in these types of systems. One
  usually measures a membrane's pore radius based on the size of a molecule it
  can reject, however it is not clear where the edges of the pores are and what
  size molecule would fit through. We leave these investigations for a future
  study.

### 2018-01-19 Ben_Manuscripts/structure_paper/supplemental.tex
More analysis scripts and figures

  \color{red}{This is an outline of this section and will be reworded for clarity}
  \begin{itemize}
  	\item We are interested in 5 pore-to-pore distances which should all
	be equal in a perfect hexagonal array, however only 4 distances are independent
	% can visualize this in supplemental material. Will make a lot of things more
	clear below.
	\item Each pore spacing has its own trajectory of spacing vs. time.
	Using data collected after the system is equilibrated, we calculate how long it
	takes for the data in each of the 5 trajetories to become uncorrelated using
	pymbar.timeseries.integratedAutocorrelationTime() % reference
	\item We break the full trajectories down into sub-trajectories based
	on the maximum autocorrelation time of those found in the previous step.
	\item For each bootstrap trial, we recreate an equilibrium trajectory
	by randomly sampling pore spacings from the sub-trajectories 
	\item We get an average value for each pore spacing by finding the
	mean of the bootstrapped data
	\item We calculate the overall average as the mean of all
	bootstrapped pore spacings
	\item The uncertainty for each pore spacing is calculated as
	$\dfrac{<x> - x}{4}$ where $<x>$ is the average spacing from the bootstrap
	trial and $x$ is the average value of one of the pore spacings.
	\item We report the mean of these uncertainties
  \end{itemize}

@@

@@

>>>>

  We are interested in 5 pore-to-pore distances which should all be
  equal in a perfect hexagonal array (Figure~\ref{fig:p2p_diagram}). Only 4 of
  the 5 distances are independent. We can calculate a trajectory of spacing
  versus time for each of the 5 distances. We calculated the average pore-to-pore
  spacing and its uncertainty according to the following procedure:
   \begin{enumerate}
	\item We calculated the time when each of the pore-to-pore distances
	were equilibrated using \texttt{pymbar.timeseries.detectEquilibration()} \\
	\cite{chodera_simple_2016,shirts_statistically_2008}. We began calculations
	after the largest of the five values.
	\item We calculated how long it takes for the data in each of the 5
	trajectories to become uncorrelated using \texttt{pymbar.timeseries.integratedAutocorrelationTime()}
	\cite{chodera_simple_2016,shirts_statistically_2008}. 
	\item We broke the full equilibrated trajectory into blocks of length $\tau$, where
	$\tau$ is the maximum of the five autocorrelation times calculated. Each block contains
	five sub-trajectories of pore-to-pore spacings.   
	\item We generate statistics using the bootstrapping technique. For
	each bootstrap trial, we reconstruct an equilibrium trajectory by randomly
	sampling from the trajectory blocks. 
	\item The average pore-to-pore distance is the mean of all pore
	spacings among all bootstrap trials.
	\item To calculate the uncertainty in pore-to-pore distance, we calculate the average 
	pore spacing for each pore over all bootstrap trials. Using the 5 average 
	pore-to-pore distances, we calculate the spread with:
	\begin{align}
	s = \sqrt{\frac{1}{4} \sum_{i=1}^5 (x_i - \overline{x})^2} 
	\end{align}
	where $\overline{x}$ is the average pore-to-pore distance.
  \end{enumerate}

  \begin{figure}
	\centering
	\includegraphics[width=0.25\textwidth]{p2p_diagram.png}
	\caption{There are five pore-to-pore distances that are equal in a perfect
		hexagonal. If each number in the diagram represents the center of a
		pore in a hexagonal unit cell, then the distance from 1 to 2, 2 to 4, 
		4 to 3, 3 to 1 and 1 to 4 should be equal. Only 4 of the distances
		are independent. For example, the distance from 1 to 4 is defined by
		the location of all other pore centers.}
	\label{fig:p2p_diagram}
  \end{figure}

@@

  \vspace{1em}
  \begingroup
	\fontsize{14pt}{12pt}\selectfont
	\textbf{Experimental correlation length of Ri-$\pi$}
  \endgroup

  \vspace{1em}
  Experimentalists did not measure the correlation length, so we estimated it from
  the raw data. We took a slice of the 2D WAXS data along the dashed line in Figure
  ~\ref{fig:waxs_dashed}. We removed background noise by subtracting
  the intensity far from the pattern (high $q$) uniformly. We used a least squares
  algorithm to fit a lorentzian curve to the data. Only values recorded above $q_z$ = 1.6 were considered for 
  the fit in order to mitigate interference from R-alkanes (Figure~\ref{fig:correlation_length_exp}).
  The full width at half maximum (FWHM) is related to the correlation length, L, by
  the relationship: $ L = \frac{\pi}{FWHM} $. The error in the value is calculated
  as the square root of the diagonal entry of the covariance matrix calculated 
  % need citations and more on lorentzian parameters
  \begin{figure}[h]
  \centering
  \begin{subfigure}{0.45\textwidth}	
  	\includegraphics[width=\textwidth]{waxs_dashed.png}
	\caption{}~\label{fig:waxs_dashed}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
	\includegraphics[width=\textwidth]{Correlation_length_exp.png}
	\caption{}~\label{fig:correlation_length_exp}
  \end{subfigure}
  \caption{}~\label{fig:correlation}
  \end{figure}

@@

  \begin{figure}[ht]

  \begin{subfigure}{\linewidth}
        \centering
        \begin{subfigure}{0.45\linewidth}
                \centering
                \includegraphics[width=\linewidth]{layered_nematic_order.png}
                \caption{Sandwiched, Ordered}~\label{fig:sandwich_nematic}
        \end{subfigure}%
        \begin{subfigure}{0.45\linewidth}
                \centering
                \includegraphics[width=\linewidth]{offset_nematic_order.png}
                \caption{Parallel Displaced, Ordered}~\label{fig:offset_nematic}
        \end{subfigure}
        \begin{subfigure}{0.45\linewidth}
                \centering
                \includegraphics[width=\linewidth]{disorder_sandwich_nematic_order.png}
                \caption{Sandwiched, Disordered}~\label{fig:disorder_sandwich_nematic}
        \end{subfigure}%
        \begin{subfigure}{0.45\linewidth}
                \centering
                \includegraphics[width=\linewidth]{disorder_offset_nematic_order.png}
                \caption{Parallel Displaced, Disordered}~\label{fig:disorder_offset_nematic}
        \end{subfigure}

  \end{subfigure}
  \caption{}~\label{fig:nematic_distribution}
  \end{figure}
